---
title: "test impulse genes in COPDGene RNAseq"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r}
library(data.table)
library(dplyr)
library(tidyr)
library(tableone)
library(stringr)
library(MASS)
library(ggplot2)
library(glmnet)
library(pROC)
library(ResourceSelection)
library(caret)
library(ggpubr)
library(survival)
library(survminer)
library(coin)

options(bitmapType="cairo")
```

# functions
```{r}
call_model <- function(df, outcome_name, top_genes, covars, adjustment=NULL, model_type) {
  results <- data.frame(gene = character(), beta = numeric(), ci_lower = numeric(), ci_upper = numeric(), beta_ci = character(), pval = numeric(), stringsAsFactors = FALSE)

  for (gene in top_genes) {
    gene <- ifelse(grepl("\\|", gene), paste0("`", gene, "`"), gene)

    formula_components <- c(gene, covars, adjustment)
    formula_str <- paste0(outcome_name, " ~ ", paste0(formula_components, collapse = " + "))
    
    tryCatch({
      if (model_type == "nb") {
        model <- glm.nb(as.formula(formula_str), data = df)
      } else if (model_type == "binomial") {
        model <- glm(as.formula(formula_str), data = df, family = binomial())
      }
        else if (model_type == "lm") {
        model <- lm(as.formula(formula_str), data = df)
      }
  

    beta <- coef(model)[gene]
    ci <- confint(model, gene, level = 0.95)
    beta_ci <- sprintf("%.3f (%.3f, %.3f)", beta, ci[1], ci[2])
    pval <- summary(model)$coefficients[gene, 4]
    results <- rbind(results, data.frame(gene = gene, beta = beta, ci_lower = ci[1], ci_upper = ci[2], beta_ci = beta_ci, pval = pval))
    }, error = function(e) {
      message(sprintf("Skipping gene %s due to error: %s", gene, e$message))
    })
  }
  return(results)
}
```


```{r}
get_ntile_summary <- function(df, outcome_name, score, covars, adjustment=NULL, model_type=NULL, ntile) {
  # create ntile column for score
  df <- df %>%
    mutate(score_ntile = ntile(.data[[score]], ntile))
 
  ntile_results <- data.frame(ntile = numeric(),
                             OR = numeric(),
                             OR_CI_lower = numeric(),
                             OR_CI_upper = numeric(), 
                             beta = numeric(),
                             beta_ci_lower = numeric(),
                             beta_ci_upper = numeric(),
                             P = numeric(),
                             N = numeric(),
                             Mean = numeric(),
                             Sd = numeric())
                              
  for (cur_ntile in 1:ntile) {
    df$cur_ntile <- ifelse(df$score_ntile == 1, 0, ifelse(df$score_ntile == cur_ntile, 1, NA))
    if (cur_ntile == 1) {
      ntile_results <- rbind(ntile_results,
                                data.frame(ntile = cur_ntile,
                                           OR = 1,
                                           CI_lower = NA,
                                           CI_upper = NA,
                                           beta = 0,
                                           beta_ci_lower = NA,
                                           beta_ci_upper = NA,
                                           P = NA,
                                           N = sum(df$score_ntile == 1, na.rm = TRUE),
                                           Mean = mean(df[df$score_ntile == 1, score], na.rm = TRUE),
                                           Sd = sd(df[df$score_ntile == 1, score], na.rm = TRUE)))
    } else {
      model_table <- call_model(df[!is.na(df$cur_ntile), ], outcome_name, top_genes = "cur_ntile", covars = covars, adjustment = adjustment, model_type = model_type)

      ntile_results <- rbind(ntile_results,
                                data.frame(ntile = cur_ntile,
                                           OR = exp(model_table$beta),
                                           CI_lower = exp(model_table$ci_lower),
                                           CI_upper = exp(model_table$ci_upper),
                                           beta = model_table$beta,
                                           beta_ci_lower = model_table$ci_lower,
                                           beta_ci_upper = model_table$ci_upper,
                                           P = model_table$pval,
                                           N = sum(df$cur_ntile == 1, na.rm = TRUE),
                                           Mean = mean(df[df$cur_ntile == 1, score], na.rm = TRUE),
                                           Sd = sd(df[df$cur_ntile == 1, score], na.rm = TRUE)))
    }
  }
  
  return(ntile_results)
}

```

```{r}
get_stratified_model <- function(df, outcome_name, ntile_var, predictor, covars=NULL, adjustment=NULL, ntile=4) {
  
  df <- df %>%
    mutate(score_ntile = ntile(.data[[ntile_var]], ntile))
  
  results <- data.frame()
  
  for (cur_ntile in 1:ntile) {
    
    df_sub <- df %>%
      filter(score_ntile == cur_ntile)
    
    model_table <- call_model(
      df = df_sub,
      outcome_name = outcome_name,
      top_genes = predictor,
      covars = covars,
      adjustment = adjustment,
      model_type = "binomial"
    )
    
    results <- rbind(results,
      data.frame(
        ntile = cur_ntile,
        predictor = predictor,
        OR = exp(model_table$beta),
        CI_lower = exp(model_table$ci_lower),
        CI_upper = exp(model_table$ci_upper),
        P = model_table$pval,
        N = nrow(df_sub)
      )
    )
  }
  
  return(results)
}
```

```{r}
call_coxph <- function(df, time, event, predictor, adjustment_vars) {
  
  surv_obj <- Surv(time = df[[time]], event = df[[event]])
  
  formula_components <- c(predictor, adjustment_vars)
  formula <- as.formula(paste0("surv_obj ~ ", paste(formula_components, collapse = " + ")))
  
  model <- coxph(formula, data = df)
  
  ph_test <- cox.zph(model)

  pred <- predict(model, type = "risk")
  c_stat <- summary(model)$concordance["C"]
  c_stat_se <- summary(model)$concordance["se(C)"]

  summ <- summary(model)$coefficients
  cis <- confint(model)
  
  beta <- summ[predictor, "coef"]
  se <- summ[predictor, "se(coef)"]
  pval <- summ[predictor, "Pr(>|z|)"]
  ci <- cis[predictor, ]
    
  hr <- exp(beta)
  hr_ci <- sprintf("%.3f (%.3f, %.3f)", hr, exp(ci[1]), exp(ci[2]))
    
  ph_pval <- if (predictor %in% rownames(ph_test$table)) ph_test$table[predictor, "p"] else NA
    
  result_df <- data.frame(predictor = predictor, hr = hr, 
             ci_lower = exp(ci[1]), ci_upper = exp(ci[2]), 
             hr_ci = hr_ci, se = se, pval = pval, n_samples = model$n, 
             n_events = summary(model)$nevent,
             ph_pval = ph_pval, c_stat = c_stat, c_stat_se = c_stat_se)

  return(list(result=result_df, ph_test=ph_test, model=model))
}
```

# load data
```{r}
rnaseq_all <- fread("path/to/counts_data.tsv", data.table = F)
nrow(rnaseq_all)
# remove zero variance
gene_vars <- apply(rnaseq_all[,-1], 1, var)
rnaseq_all <- rnaseq_all[gene_vars > 0, ]
nrow(rnaseq_all)

rnaseq_all$GENEID <- sub("\\..*", "", rnaseq_all$GENEID)
```

```{r}
load("path/to/pheno_data.Rdata")

# ChanRev12 
cat("in COPDGene pheno P2: ", nrow(pheno_p2), "\n")

lfu <- fread("path/to/followup_data.txt", data.table = F)
lfu_exac <- lfu %>% 
  filter(!is.na(days_since_P2)) %>%
  group_by(sid) %>%
  summarise(total_days_since_P2 = max(days_since_P2, na.rm = TRUE))


surv_df <- lfu %>%
  filter(!is.na(days_since_P2)) %>%
  group_by(sid) %>%
  summarise(
    time_to_first_exac = ifelse(
      any(Net_Exacerbations > 0),
      min(days_since_P2[Net_Exacerbations > 0], na.rm = TRUE),
      max(days_since_P2, na.rm = TRUE)   # if no exacerbation time, is the follow up time
    ))

pheno_p2 <- pheno_p2 %>% dplyr::select(-total_days_since_P2)
pheno_p2 <- merge(pheno_p2, lfu_exac, by="sid") 
# ChanRev12 
cat("with lfu: ", nrow(pheno_p2), "\n")
pheno_p2 <- merge(pheno_p2, surv_df, by="sid") 
cat("with survival data: ", nrow(pheno_p2), "\n")

sapply(pheno_p2[c("Exacerbation_Frequency_P2", "ATS_PackYears_P2", "total_days_since_P2")], function(x) sum(is.na(x)))
pheno_p2 <- pheno_p2[complete.cases(pheno_p2[, c("Exacerbation_Frequency_P2", "ATS_PackYears_P2", "total_days_since_P2")]), ]
# ChanRev12 
cat("remove missingness in covars: ", nrow(pheno_p2), "\n")

pheno_p2$copd <- ifelse(pheno_p2$FEV1_FVC_post_P2 < 0.7 & pheno_p2$fev1_pp_gli_global_P2 < 80, 1,
                  ifelse(pheno_p2$FEV1_FVC_post_P2 >= 0.7 & pheno_p2$fev1_pp_gli_global_P2 >= 80, 0, NA))

pheno_copd <- pheno_p2 %>% filter(copd == 1)
# ChanRev12 
cat("COPD cases: ", nrow(pheno_copd), "\n")

# fill pack years with 0 if smoking is 0
pheno_copd$ATS_PackYears_P2[is.na(pheno_copd$ATS_PackYears_P2) & pheno_copd$SmokCigNow_P2 == 0] <- 0

pheno_copd <- pheno_copd %>% mutate(
    exacerbation_rate = total_Net_Exacerbations / total_days_since_P2 * 365,
    has_exacerbation = if_else(total_Net_Exacerbations >= 1, 1, 0),
    exac_0or1vs2ormore = if_else(total_Net_Exacerbations >= 2, 1, 0),
    exac_0vs2ormore = case_when(total_Net_Exacerbations >= 2 ~ 1, total_Net_Exacerbations == 0 ~ 0, TRUE ~ NA_real_),
    has_prior_exac = if_else(Exacerbation_Frequency_P2 > 0, 1, 0)
  )
```

```{r}
pheno_copd <- pheno_copd %>%
  mutate(
    died = mortality_survival_vital_status == 1,
    has_exac = has_exacerbation == 1,
    
    composite_event = as.integer(died | has_exac),

    composite_time = case_when(
      died & has_exac ~ pmin(total_days_since_P2, time_to_first_exac),
      died ~ total_days_since_P2,
      has_exac ~ time_to_first_exac,
      TRUE ~ total_days_since_P2   # censored at end of follow up
    )
  )
```


```{r}
# from train_lasso_score.Rmd
module_weights_df <- read.csv("../data/impulse_gene_score_WGCNA_AMEND_adalasso.csv")
nrow(module_weights_df)
module_weights_df <- module_weights_df %>% 
  rename("Gene_ID"="feature") 
```

```{r}
rnaseq_module <- rnaseq_all %>% filter(GENEID %in% module_weights_df$Gene_ID)

GENEID_module <- rnaseq_module[, 1]
rnaseq_mtx_module <- rnaseq_module[, -1]
rnaseq_transpose_module <- as.data.frame(t(rnaseq_mtx_module))
colnames(rnaseq_transpose_module) <- GENEID_module

rnaseq_transpose_module$sid <- rownames(rnaseq_transpose_module)
rownames(rnaseq_transpose_module) <- NULL

rnaseq_transpose_module <- rnaseq_transpose_module %>%
  mutate(across(-sid, ~ log2(.x + 1)))

all(setdiff(colnames(rnaseq_transpose_module), "sid") %in% module_weights_df$Gene_ID)

df2 <- merge(rnaseq_transpose_module, pheno_copd, by="sid")

# for regressing out
df2$has_exacerbation <- as.factor(df2$has_exacerbation)
df2$gender <- as.factor(df2$gender)
df2$race <- as.factor(df2$race)
# ChanRev12 
cat("with RNAseq data: ", nrow(df2), "\n")
```

# WGCNA module genes (28)
```{r}
DT::datatable(module_weights_df %>%
                dplyr::select(Gene_ID, Gene_Name, lambda.min_ada_subj_fold) %>%
                arrange(desc(abs(lambda.min_ada_subj_fold))))
```

## adalasso from PeriCOPD (6)
```{r}
score_col <- "lambda.min_ada_subj_fold"
score_name <- "module_score_ada_subj_fold"

selected_df <- module_weights_df %>%  filter(.data[[score_col]] != 0)

selected_genes <- selected_df$Gene_ID
beta_original <- selected_df[[score_col]]

df3 <- df2
# regress out in df3 for models without train test split
regress_vars <- c("Age_P2","gender")

for(gene in selected_genes){
      df3[,gene] <- as.numeric(resid(lm(as.formula(paste0(gene,"~",paste0(regress_vars,collapse = "+"))),data=df3)))
    }

# compute score
X <- as.matrix(df3[, selected_genes])
new_score <- as.vector(X %*% beta_original)

df3[[score_name]] <- scale(new_score)

res_has_exac <- call_model(df=df3, outcome_name="has_exacerbation", top_genes=score_name, covars=c("Exacerbation_Frequency_P2", "ATS_PackYears_P2"), adjustment="offset(log(total_days_since_P2/365))", model_type="binomial")

res_exac_0vs2ormore <- call_model(df=df3, outcome_name="exac_0vs2ormore", top_genes=score_name, covars=c("Exacerbation_Frequency_P2", "ATS_PackYears_P2"), adjustment="offset(log(total_days_since_P2/365))", model_type="binomial")

res_exac_0or1vs2ormore <- call_model(df=df3, outcome_name="exac_0or1vs2ormore", top_genes=score_name, covars=c("Exacerbation_Frequency_P2", "ATS_PackYears_P2"), adjustment="offset(log(total_days_since_P2/365))", model_type="binomial")

res_module_ada <- rbind(res_has_exac, res_exac_0vs2ormore, res_exac_0or1vs2ormore) %>%
  mutate(outcome = c("0 vs. 1 or more", "0 vs. 2 or more", "0 or 1 vs. 2 or more")) %>%
  rename(score = gene)
DT::datatable(res_module_ada[, c("score", "outcome", "beta_ci", "pval")], rownames = FALSE)
```

```{r}
formula_components <- c("module_score_ada_subj_fold", "Exacerbation_Frequency_P2", "ATS_PackYears_P2", "offset(log(total_days_since_P2/365))")
formula_str <- paste0("has_exacerbation ~ ", paste0(formula_components, collapse = " + "))
model_module_no_reweight <- glm(as.formula(formula_str), data = df3, family = binomial())
# summary(model_module_no_reweight)

df3$pred_prob <- predict(model_module_no_reweight, newdata = df3, type = "response")
roc_module_no_reweight <- roc(response  = df3$has_exacerbation, predictor = df3$pred_prob)
auc(roc_module_no_reweight)
```

## individual genes
```{r}
res_indiv_genes_w_covars <- call_model(df=df3, outcome_name="has_exacerbation", top_genes=selected_genes, covars=c("Exacerbation_Frequency_P2", "ATS_PackYears_P2"), adjustment="offset(log(total_days_since_P2/365))", model_type="binomial")

res_indiv_genes_w_covars <- res_indiv_genes_w_covars %>% merge(module_weights_df %>% dplyr::select(-lambda.min_ada_adj_subj), by.x = "gene", by.y = "Gene_ID")

DT::datatable(res_indiv_genes_w_covars[, c("Gene_Name", "lambda.min_ada_subj_fold", "beta_ci", "pval")] %>% mutate_if(is.numeric, signif, digits=3), rownames = FALSE)
```
```{r}
# get sign of individual genes ~ has_exac, no covars
res_indiv_genes <- call_model(df=df3, outcome_name="has_exacerbation", top_genes=selected_genes, covars=NULL, adjustment=NULL, model_type="binomial")

res_indiv_genes <- res_indiv_genes %>% merge(module_weights_df %>% dplyr::select(-lambda.min_ada_adj_subj), by.x = "gene", by.y = "Gene_ID")

# flip the direction of the score to be the same direction as the genes ~ has_exac
res_indiv_genes$lambda.min_ada_subj_fold <- abs(res_indiv_genes$lambda.min_ada_subj_fold) * sign(res_indiv_genes$beta)

DT::datatable(res_indiv_genes[, c("Gene_Name", "lambda.min_ada_subj_fold", "beta", "pval")] %>% mutate_if(is.numeric, signif, digits=3), rownames = FALSE)
```

## orient direction of weight (sig)
```{r}
selected_genes <- res_indiv_genes$gene
beta_flip <- res_indiv_genes$lambda.min_ada_subj_fold

X <- as.matrix(df3[, selected_genes])
new_score <- as.vector(X %*% beta_flip)

# score_name <- "module_score_oriented"
df3$module_score_oriented <- scale(new_score)

# oriented score and clinical
formula_components <- c("module_score_oriented", "Exacerbation_Frequency_P2", "ATS_PackYears_P2", "offset(log(total_days_since_P2/365))")
formula_str <- paste0("has_exacerbation ~ ", paste0(formula_components, collapse = " + "))
model_oriented_clinical <- glm(as.formula(formula_str), data = df3, family = binomial())

beta_all <- coef(model_oriented_clinical)
ci_all <- confint(model_oriented_clinical, level = 0.95)
pval_all <- summary(model_oriented_clinical)$coefficients[, 4]

module_flip_table <- data.frame(
  predictor = names(beta_all),
  beta = beta_all,
  ci_lower = ci_all[, 1],
  ci_upper = ci_all[, 2],
  pval = pval_all,
  row.names = NULL) %>%
  mutate(or = exp(beta), ci_lower = exp(ci_lower), ci_upper = exp(ci_upper), or_ci = sprintf("%.3f (%.3f, %.3f)", or, ci_lower, ci_upper)) %>%
  mutate_if(is.numeric, signif, digits=3) %>%
  filter(!predictor %in% c("(Intercept)")) %>%
  # mutate(predictor = c("NERD score", "pack years", "piror exacerbation")) %>%
  dplyr::select(predictor, or_ci, pval)

DT::datatable(module_flip_table)
```


# reweight (sig)
```{r eval=FALSE}
score_cols <- "lambda.min_ada_subj_fold"
score_names <- "nerd_score"

set_seeds <- 101:200

res_module_ada_reweight_list <- vector("list", length(set_seeds))

for (i in seq_along(set_seeds)) {
  set.seed(set_seeds[i])
  train_indices <- createDataPartition(df2$has_exacerbation, p = 0.5, list = FALSE)
  df_train <- df2[train_indices, ]
  df_test  <- df2[-train_indices, ]
  
  reweighted_scores <- list()
  for (j in seq_along(score_cols)) {
    score_col <- score_cols[j]
    score_name <- score_names[j]
  
    selected_df <- module_weights_df %>%
      filter(.data[[score_col]] != 0)
    selected_genes <- selected_df$Gene_ID
    
    for(gene in selected_genes){
      df_train[,gene] <- as.numeric(resid(lm(as.formula(paste0(gene,"~",paste0(regress_vars,collapse = "+"))),data=df_train)))
      df_test[,gene] <- as.numeric(resid(lm(as.formula(paste0(gene,"~",paste0(regress_vars,collapse = "+"))),data=df_test)))
    }

    all_terms <- c(selected_genes, "Exacerbation_Frequency_P2", "ATS_PackYears_P2", "offset(log(total_days_since_P2 / 365))")
      
    formula_str <- paste("has_exacerbation ~", paste(all_terms, collapse = " + "))
    refit_model <- glm(as.formula(formula_str), data = df_train, family = binomial())
  
    beta_reweighted <- coef(refit_model)[selected_genes]
  
    X_test <- as.matrix(df_test[, selected_genes])
    new_score <- as.vector(X_test %*% beta_reweighted)

    df_test[[score_name]] <- scale(new_score)

    res <- call_model(df=df_test, outcome_name="has_exacerbation", top_genes=score_name, covars=c("Exacerbation_Frequency_P2", "ATS_PackYears_P2"), adjustment="offset(log(total_days_since_P2/365))", model_type="binomial")
    
    res$seed <- set_seeds[i]
    res_module_ada_reweight_list[[length(res_module_ada_reweight_list) + 1]] <- res
    
  }
}

res_module_ada_reweight <- do.call(rbind, res_module_ada_reweight_list) %>%
  rename(score = gene)

write.table(res_module_ada_reweight, "../data/cg_RNAseq_reweighted_WGCNA_AMEND_scores_smk_exac_only_regress_age_sex_test_seeds_sumtab_2.txt", row.names = F, quote = F, sep = "\t")
```


```{r}
res_module_ada_reweight <- read.table("../data/cg_RNAseq_reweighted_WGCNA_AMEND_scores_smk_exac_only_regress_age_sex_test_seeds_sumtab_2.txt", header = T, sep = "\t")

sig_counts_module_ada_reweight <- res_module_ada_reweight %>%
  group_by(score) %>%
  summarise(n_significant = sum(pval < 0.05, na.rm = TRUE))

sig_counts_module_ada_reweight
```

```{r}
best_module_ada_reweight <- res_module_ada_reweight %>% 
  filter(score == "nerd_score" & pval < 0.05) %>%
  arrange(pval) %>%
  slice(1) # min p val

best_module_ada_reweight
```


# NERD score
## train test
```{r}
set.seed(138)
train_indices <- createDataPartition(df2$has_exacerbation, p = 0.5, list = FALSE)
df_train_final <- df2[train_indices, ]
df_test_final  <- df2[-train_indices, ]
score_col <- "lambda.min_ada_subj_fold"

selected_df <- module_weights_df %>%
  filter(.data[[score_col]] != 0)
selected_genes <- selected_df$Gene_ID

gene_resid_models <- list()
for(gene in selected_genes){
  gene_resid_models[[gene]] <- lm(as.formula(paste0(gene,"~",paste0(regress_vars,collapse = "+"))),data=df_train_final)
  # apply train model to both train and test
  df_train_final[, gene] <- as.numeric(resid(gene_resid_models[[gene]]))
  df_test_final[, gene]  <- df_test_final[, gene] -
    as.numeric(predict(gene_resid_models[[gene]], newdata = df_test_final))
}

all_terms <- c(selected_genes, "Exacerbation_Frequency_P2", "ATS_PackYears_P2", "offset(log(total_days_since_P2 / 365))")

formula_str <- paste("has_exacerbation ~", paste(all_terms, collapse = " + "))
refit_model <- glm(as.formula(formula_str), data = df_train_final, family = binomial())

beta_reweighted <- coef(refit_model)[selected_genes]

X_test <- as.matrix(df_test_final[, selected_genes])
df_test_final$nerd_score <- as.vector(X_test %*% beta_reweighted)

q4_cutoff <- quantile(df_test_final$nerd_score, 0.75, na.rm = TRUE)
summary(df_test_final$nerd_score)
sd(df_test_final$nerd_score)
cat("Q4 nerd score cutoff before scaling:", q4_cutoff)

df_test_final$nerd_score <- scale(df_test_final$nerd_score)
```

```{r}
# regress out age and sex from prior exac
exac_regress_age_sex <- lm(as.formula(paste0("Exacerbation_Frequency_P2 ~", paste0(regress_vars, collapse="+"))),
                            data = df_train_final)
df_train_final$prior_exac_resid <- as.numeric(resid(exac_regress_age_sex))
df_test_final$prior_exac_resid  <- df_test_final$Exacerbation_Frequency_P2 -
  as.numeric(predict(exac_regress_age_sex, newdata = df_test_final))
coef(exac_regress_age_sex)

# regress out age and sex from pack years
py_regress_age_sex <- lm(as.formula(paste0("ATS_PackYears_P2 ~", paste0(regress_vars, collapse="+"))),
                            data = df_train_final)
df_train_final$py_resid <- as.numeric(resid(py_regress_age_sex))
df_test_final$py_resid  <- df_test_final$ATS_PackYears_P2 -
  as.numeric(predict(py_regress_age_sex, newdata = df_test_final))

coef(py_regress_age_sex)
```

```{r}
resid_coefs <- t(sapply(selected_genes, function(gene) {
  coef(gene_resid_models[[gene]])
}))

age_weight <- sum(-beta_reweighted * resid_coefs[, "Age_P2"])
sex_weight <- sum(-beta_reweighted * resid_coefs[, "gender2"])
intercept_adj <- sum(-beta_reweighted * resid_coefs[, "(Intercept)"])

# score <- sum(beta_reweighted * raw_genes) + age_weight * Age + sex_weight * Sex + intercept_adj
gene_formula_terms <- paste0(beta_reweighted, " * ", names(beta_reweighted))

formula_text <- paste(
  "score =",
  paste(gene_formula_terms, collapse = " + "),
  "+", age_weight, "* age",
  "+", sex_weight, "* sex",
  "+", intercept_adj
)

formula_text
```

```{r}
module_6_gene_score <- module_weights_df %>%
  filter(lambda.min_ada_subj_fold != 0) %>%
  rename(score_weights = lambda.min_ada_subj_fold) %>%
  dplyr::select(Gene_Name, Gene_ID, score_weights) %>%
  arrange(desc(abs(score_weights))) 

beta_reweighted_df <- as.data.frame(beta_reweighted)
beta_reweighted_df$Gene_ID <- rownames(beta_reweighted_df)
module_6_gene_score <- merge(module_6_gene_score, beta_reweighted_df, by = "Gene_ID")

formula_table <- data.frame(
  Term = c(names(beta_reweighted), "age", "sex", "intercept"),
  Coefficient = c(beta_reweighted, age_weight, sex_weight, intercept_adj)
)

module_score_w_age_sex <- merge(formula_table, module_6_gene_score, by.x = "Term",  by.y = "Gene_ID", all.x = T) %>%
  mutate(
    Predictor = case_when(Term %in% c("age", "sex", "intercept") ~ Term, TRUE ~ Gene_Name),
    Term = case_when(Term %in% c("age", "sex", "intercept") ~ NA_character_, TRUE ~ Term)) %>%
  arrange(match(Predictor, c(sort(Gene_Name), "age", "sex", "intercept"))) %>%
  rename(COPDGene_reweighted = Coefficient, ENSG_ID = Term, PeriCOPD_weights = score_weights) %>%
  dplyr::select(Predictor, COPDGene_reweighted, ENSG_ID, PeriCOPD_weights)
write.table(module_score_w_age_sex, "../data/impulse_gene_score_WGCNA_AMEND_adalasso_subj_fold_selected_w_age_sex.txt", sep = "\t", quote = F, row.names = F)
# ChanRev16 
# score training in train_lasso_score.Rmd
DT::datatable(module_score_w_age_sex, rownames = F)
```

```{r}
# sanity check
df_test_raw <- df2[-train_indices, ]
X_raw <- as.matrix(df_test_raw[, selected_genes])
df_test_final$score_explicit <- as.vector(X_raw %*% beta_reweighted) +
  age_weight * df_test_raw$Age_P2 +
  sex_weight * as.numeric(df_test_raw$gender) +
  intercept_adj
df_test_final$score_explicit <- scale(df_test_final$score_explicit)

# same score from formula and from train test split, except rounding
max(abs(df_test_final$nerd_score - df_test_final$score_explicit))
```

## multivariable sumtab
```{r message=FALSE, warning=FALSE}
formula_str <- paste0("has_exacerbation ~ ", paste0(c("nerd_score", "Exacerbation_Frequency_P2", "ATS_PackYears_P2", "SmokCigNow_P2", "race", "offset(log(total_days_since_P2/365))"), collapse = " + "))
model_nerd_exac_py_smk_race <- glm(as.formula(formula_str), data = df_test_final, family = binomial())

beta_all <- coef(model_nerd_exac_py_smk_race)
ci_all <- confint(model_nerd_exac_py_smk_race, level = 0.95)
pval_all <- summary(model_nerd_exac_py_smk_race)$coefficients[, 4]

nerd_exac_py_smk_race_table <- data.frame(
  predictor = names(beta_all),
  beta = beta_all,
  ci_lower = ci_all[, 1],
  ci_upper = ci_all[, 2],
  pval = pval_all,
  row.names = NULL) %>%
  mutate(or = exp(beta), ci_lower = exp(ci_lower), ci_upper = exp(ci_upper), or_ci = sprintf("%.2f (%.2f, %.2f)", or, ci_lower, ci_upper)) %>%
  mutate(pval = ifelse(pval < 0.001, "<.001",
                ifelse(pval < 0.01, sub("^0", "", sprintf("%.3f", pval)),
                                    sub("^0", "", sprintf("%.2f", pval))))) %>%
  filter(!predictor %in% c("(Intercept)")) %>%
  mutate(predictor = c("NERD score", "Prior exacerbation", "Pack years", "Current smoking", "Race")) %>%
  dplyr::select(predictor, or_ci, pval)

nerd_exac_py_smk_table_ec <- read.table("../data/multivariable_model_sumtab_eclipse_adj_smk_status.txt", sep = "\t", header = T)
nerd_exac_py_smk_race_merged <- merge(nerd_exac_py_smk_race_table, nerd_exac_py_smk_table_ec, by="predictor", all.x=T, suffix = c("_COPDGene", "_ECLIPSE")) %>% 
  arrange(match(predictor, c("NERD score", "Prior exacerbation", "Pack years", "Current smoking", "Race")))
write.table(nerd_exac_py_smk_race_merged, "../data/multivariable_model_sumtab_adj_smk_race.txt", row.names = F, quote = F, sep = "\t")
# ChanRev20
# no time offset to this table because coefficient is permanently set to 1
DT::datatable(nerd_exac_py_smk_race_merged, rownames = F)
```

```{r}
formula_str <- paste0("has_exacerbation ~ ", paste0(c("nerd_score", "Exacerbation_Frequency_P2", "ATS_PackYears_P2", "SmokCigNow_P2", "race", "log(total_days_since_P2/365)"), collapse = " + "))
model_nerd_exac_py_smk_race_time_no_offset <- glm(as.formula(formula_str), data = df_test_final, family = binomial())

beta_all <- coef(model_nerd_exac_py_smk_race_time_no_offset)
ci_all <- confint(model_nerd_exac_py_smk_race_time_no_offset, level = 0.95)
pval_all <- summary(model_nerd_exac_py_smk_race_time_no_offset)$coefficients[, 4]


nerd_exac_py_smk_race_time_no_offset_table <- data.frame(
  predictor = names(beta_all),
  beta = beta_all,
  ci_lower = ci_all[, 1],
  ci_upper = ci_all[, 2],
  pval = pval_all,
  row.names = NULL) %>%
  mutate(or = exp(beta), ci_lower = exp(ci_lower), ci_upper = exp(ci_upper), or_ci = sprintf("%.3f (%.3f, %.3f)", or, ci_lower, ci_upper)) %>%
  mutate_if(is.numeric, signif, digits=3) %>%
  filter(!predictor %in% c("(Intercept)")) %>%
  mutate(predictor = c("NERD score", "Prior exacerbation", "Pack years", "Current smoking", "Race", "Log time")) %>%
  dplyr::select(predictor, or_ci, pval)

DT::datatable(nerd_exac_py_smk_race_time_no_offset_table, rownames = F)
```

### adjust for cell count
```{r}
cell_df <- read.table("path/to/cell_count_data.txt", sep="\t", header = TRUE, quote = "", fill = TRUE)[, c("sid", "Phase_study", "eosinphl_pct_calc", "neutrophl_pct_calc", "lymphcyt_pct_calc", "monocyt_pct_calc")]
cell_df_p2 <- cell_df %>% filter(Phase_study == 2)

df_test_final_cell <- merge(df_test_final, cell_df_p2, all.x=TRUE, by="sid")

cell_vars <- c("eosinphl_pct_calc", "neutrophl_pct_calc", "lymphcyt_pct_calc", "monocyt_pct_calc")
colSums(is.na(df_test_final_cell[cell_vars]))
```

```{r}
# print out cell
formula_str <- paste0("has_exacerbation ~ ", paste0(c("nerd_score", "Exacerbation_Frequency_P2", "ATS_PackYears_P2", "SmokCigNow_P2", "race", cell_vars,
                                                      "offset(log(total_days_since_P2/365))"), collapse = " + "))
model_cell_adj <- glm(as.formula(formula_str), data = df_test_final_cell, family = binomial())

beta_cell <- coef(model_cell_adj)
ci_cell <- confint(model_cell_adj, level = 0.95)
pval_cell <- summary(model_cell_adj)$coefficients[, 4]

nerd_exac_py_smk_race_cell_table <- data.frame(
  predictor = names(beta_cell),
  beta = beta_cell,
  ci_lower = ci_cell[, 1],
  ci_upper = ci_cell[, 2],
  pval = pval_cell,
  row.names = NULL) %>%
  mutate(or = exp(beta), ci_lower = exp(ci_lower), ci_upper = exp(ci_upper), or_ci = sprintf("%.2f (%.2f, %.2f)", or, ci_lower, ci_upper)) %>%
  mutate_if(is.numeric, signif, digits=2) %>%
  filter(!predictor %in% c("(Intercept)")) %>%
  mutate(predictor = c("NERD score", "Prior exacerbation", "Pack years", "Current smoking", "Race",
                       "Eosinophils (%)", "Neutrophils (%)", "Lymphocytes (%)", "Monocytes (%)")) %>%
  dplyr::select(predictor, or_ci, pval)

adj_cell_sumtab <- merge(nerd_exac_py_smk_race_table, nerd_exac_py_smk_race_cell_table,
                         by = "predictor", all = TRUE,
                         suffixes = c("_without_cell_count", "_with_cell_count")) %>%
  arrange(match(predictor, nerd_exac_py_smk_race_cell_table$predictor))

DT::datatable(adj_cell_sumtab, rownames = FALSE, options = list(scrollX = TRUE))
write.table(adj_cell_sumtab, "../data/multivariable_model_sumtab_adj_cell_count_all_vars.txt",
            sep = "\t", row.names = FALSE, quote = FALSE)
```


## table one
```{r}
allVars <- c("Age_P2", "gender", "race", "SmokCigNow_P2", "ATS_PackYears_P2", "fev1_pp_gli_global_P2", "FEV1_FVC_post_P2", "has_prior_exac", "Exacerbation_Frequency_P2", "total_days_since_P2", "has_exacerbation", "exacerbation_rate", "copd")
catVars <- c("gender", "race", "SmokCigNow_P2", "has_prior_exac", "has_exacerbation", "copd")
table1 <- CreateTableOne(data = df2, vars = allVars, factorVars = catVars)
table1_output <- print(table1, showAllLevels = TRUE, nonnormal = c("total_days_since_P2", "fev1_pp_gli_global_P2"))
```

```{r}
table1_train <- CreateTableOne(data = df_train_final, vars = allVars, factorVars = catVars)
table1_train_output <- print(table1_train, showAllLevels = TRUE, nonnormal = c("total_days_since_P2", "fev1_pp_gli_global_P2"))

table1_test <- CreateTableOne(data = df_test_final, vars = allVars, factorVars = catVars)
table1_test_output <- print(table1_test, showAllLevels = TRUE, nonnormal = c("total_days_since_P2", "fev1_pp_gli_global_P2"))

table1_all_df <- data.frame(Variable = rownames(table1_output), table1_output, row.names = NULL)
colnames(table1_all_df) <- c("Variable", "Level", "COPDGene full")

table1_train_df <- data.frame(Variable = rownames(table1_train_output), table1_train_output, row.names = NULL)
colnames(table1_train_df) <- c("Variable", "Level", "COPDGene training")

table1_test_df <- data.frame(Variable = rownames(table1_test_output), table1_test_output, row.names = NULL)
colnames(table1_test_df) <- c("Variable", "Level", "COPDGene testing")

table1_combined <- cbind(
  table1_all_df,
  `COPDGene training` = table1_train_df$`COPDGene training`,
  `COPDGene testing` = table1_test_df$`COPDGene testing`
)

# DT::datatable(table1_combined, rownames=F)
# write.table(table1_combined, "../data/table1_cg_RNAseq", sep = "\t", row.names = TRUE, quote = FALSE)

table1_ec <- read.table("../data/table1_eclipse.txt", sep = "\t", header = T)

# fill race2
table1_ec <- table1_ec %>% add_row(Variable = "", level = 2, ECLIPSE = "0 (0.0)", .after = 5)

table1_combined2 <- cbind(
  table1_combined,
  ECLIPSE = table1_ec$ECLIPSE
)
# ChanRev13 
DT::datatable(table1_combined2, rownames=F, options = list(pageLength = 20))
write.table(table1_combined2, "../data/table1_cg_ec_combined.txt", sep = "\t", row.names = FALSE, quote = FALSE)
```

## AUC
```{r message=FALSE, warning=FALSE}
# nerd only
model_nerd <- glm(has_exacerbation ~ nerd_score + offset(log(total_days_since_P2/365)), data = df_test_final, family = binomial())
df_test_final$pred_prob_nerd <- predict(model_nerd, type = "response")
roc_nerd <- roc(response = df_test_final$has_exacerbation, predictor = df_test_final$pred_prob_nerd)

# prior exac and nerd score
model_nerd_exac <- glm(has_exacerbation ~ nerd_score + Exacerbation_Frequency_P2 + offset(log(total_days_since_P2/365)), data = df_test_final, family = binomial())
df_test_final$pred_prob_nerd_exac <- predict(model_nerd_exac, type = "response")
roc_nerd_exac <- roc(response = df_test_final$has_exacerbation, predictor = df_test_final$pred_prob_nerd_exac)

# prior exac only
model_exac <- glm(has_exacerbation ~ Exacerbation_Frequency_P2 + offset(log(total_days_since_P2/365)), data = df_test_final, family = binomial())
df_test_final$pred_prob_exac <- predict(model_exac, type = "response")
roc_exac <- roc(response = df_test_final$has_exacerbation, predictor = df_test_final$pred_prob_exac)
```

```{r}
auc_diff_row <- function(roc_new, roc_ref) {
  d <- as.numeric(auc(roc_new) - auc(roc_ref))
  se <- sqrt(var(roc_new, method = "delong") + var(roc_ref, method = "delong") -
               2 * cov(roc_new, roc_ref, method = "delong"))
  p <- roc.test(roc_new, roc_ref, method = "delong")$p.value
  c(diff = d, lo = d - 1.96 * se, hi = d + 1.96 * se, p = p)
}
fmt_num <- function(x) sprintf("%.3f", x)
fmt_auc <- function(roc_obj) {
  ci <- as.numeric(ci.auc(roc_obj))
  sprintf("%.3f (%.3f, %.3f)", ci[2], ci[1], ci[3])
}
make_auc_table <- function(rows) {
  do.call(rbind, lapply(names(rows), function(lbl) {
    r <- rows[[lbl]]
    if (is.null(r$ref)) {
      data.frame(Model = lbl, `AUC (95% CI)` = fmt_auc(r$roc),
                 `Difference vs reference (95% CI)` = "Reference",
                 `P value` = "NA", check.names = FALSE)
    } else {
      s <- auc_diff_row(r$roc, r$ref)
      data.frame(Model = lbl, `AUC (95% CI)` = fmt_auc(r$roc),
                 `Difference vs reference (95% CI)` =
                   sprintf("%s (%s, %s)", fmt_num(s["diff"]), fmt_num(s["lo"]), fmt_num(s["hi"])),
                 `P value` = round(s["p"], 3), check.names = FALSE)
    }
  }))
}

auc_table_cg <- make_auc_table(list(
  "Prior exacerbation alone" = list(roc = roc_exac),
  "Prior exacerbation and NERD score" = list(roc = roc_nerd_exac, ref = roc_exac)
))

auc_table_cg$cohort <- "COPDGene"

auc_table_ec <- read.table("../data/auc_diff_table_ECLIPSE.txt", sep = "\t", header = T)
auc_table_ec$cohort <- "ECLIPSE"
names(auc_table_ec) <- names(auc_table_cg)

auc_table <- rbind(auc_table_cg, auc_table_ec) %>%
  relocate(cohort, .before = Model)

DT::datatable(auc_table, rownames = F)
write.table(auc_table, "../data/auc_diff_table.txt", sep = "\t", row.names = FALSE, quote = FALSE)
```

```{r}
models <- c("NERD score", "Prior Exacerbations",  "Prior Exacerbations + NERD")
rocs <- list(roc_nerd, roc_exac, roc_nerd_exac)

auc_df <- data.frame(model = models, auc = sapply(rocs, auc))

ci_mat <- t(sapply(rocs, ci.auc))
auc_df$ci_low  <- ci_mat[,1]
auc_df$ci_high <- ci_mat[,3]

auc_df$model <- factor(auc_df$model, levels = models)

auc_df_ec <- read.table("../data/auc_df_nerd_vs_smk_exac_eclipse.txt", header = T, sep = "\t") %>%
  filter(!model %in% c("Pack-years", "Pack-years + NERD"))
auc_df_merged <- merge(auc_df, auc_df_ec, by = "model", suffixes = c("_COPDGene", "_ECLIPSE"))

# ChanRev22
DT::datatable(auc_df_merged %>% mutate_if(is.numeric, signif, digits=3), rownames = F, options = list(scrollX=TRUE))
```

```{r fig.height=6, fig.width=5}
auc_long <- auc_df_merged %>%
  dplyr::select(model, auc_COPDGene, ci_low_COPDGene, ci_high_COPDGene,
         auc_ECLIPSE,  ci_low_ECLIPSE,  ci_high_ECLIPSE) %>%
  pivot_longer(cols = -model, names_to = c(".value", "cohort"), names_pattern = "(.+)_(.+)$") %>% 
  mutate(model_num = as.numeric(model), offset = ifelse(cohort == "COPDGene", -0.15, 0.15), x_pos = model_num + offset)

p <- ggplot(auc_long, aes(y = auc, color = cohort)) +

  geom_errorbar(aes(x = x_pos, ymin = ci_low, ymax = ci_high), width = 0.12, linewidth = 0.8) +
  geom_segment(aes(x = x_pos - 0.06, xend = x_pos + 0.06, y = auc, yend = auc), linewidth = 0.8) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +

  scale_x_continuous(breaks = unique(auc_long$model_num), 
                     labels = c("NERD score", "Prior Exacerbations", "Prior Exacerbations\n+ NERD")) +
  labs(y = "AUC (95% CI)", x = NULL, color = "Cohort") +

  theme_bw(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  scale_color_manual(values = c("COPDGene" = "#0033FF",
                                "ECLIPSE" = "#00CCFF")) +
  scale_y_continuous(limits = c(0.48, 0.8)) +

  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "inside",
    legend.position.inside = c(0.02, 0.98),
    legend.justification = c(0, 1),
    legend.background = element_rect(fill = "white", color = "grey70")
  )
print(p)

ggsave("../figures_final/roc_nerd_exac_forest_vertical_no_clinical.png", width = 5, height = 6, dpi = 600)
ggsave("../figures_final/roc_nerd_exac_forest_vertical_no_clinical.eps", width = 5, height = 6, device = "eps")
```

## compare with de novo score
```{r}
# score training in train_new_score_cg_RNAseq.R
de_novo_score <- read.table("../data/cgRNAseq_all_genes_new_scores_regress_age_sex.txt", header = TRUE, sep = "\t")
gene_terms <- de_novo_score %>% filter(!Term %in% c("age", "sex", "intercept"))
age_w <- de_novo_score$Coefficient[de_novo_score$Term == "age"]
sex_w <- de_novo_score$Coefficient[de_novo_score$Term == "sex"]
intercept_w <- de_novo_score$Coefficient[de_novo_score$Term == "intercept"]
# ChanRev17 
nrow(gene_terms)
```

```{r message=FALSE, warning=FALSE}
rnaseq_denovo <- rnaseq_all %>% filter(GENEID %in% gene_terms$Term)

GENEID_denovo <- rnaseq_denovo[, 1]
rnaseq_mtx_denovo <- rnaseq_denovo[, -1]
rnaseq_transpose_denovo <- as.data.frame(t(rnaseq_mtx_denovo))
colnames(rnaseq_transpose_denovo) <- GENEID_denovo

rnaseq_transpose_denovo$sid <- rownames(rnaseq_transpose_denovo)
rownames(rnaseq_transpose_denovo) <- NULL

rnaseq_transpose_denovo <- rnaseq_transpose_denovo %>%
  mutate(across(-sid, ~ log2(.x + 1)))
 
# people in test
df4 <- merge(rnaseq_transpose_denovo, df_test_final, by="sid")
X <- as.matrix(df4[, gene_terms$Term])

# compute score
df4$de_novo_score <- as.vector(X %*% gene_terms$Coefficient) +
  age_w * df4$Age_P2 +
  sex_w * as.numeric(df4$gender) +
  intercept_w
all(df4$sid == df_test_final$sid)
df_test_final$de_novo_score <- scale(df4$de_novo_score)


# de novo only
model_de_novo <- glm(has_exacerbation ~ de_novo_score + offset(log(total_days_since_P2/365)), data = df_test_final, family = binomial())
df_test_final$pred_prob_de_novo <- predict(model_de_novo, type = "response")
roc_de_novo <- roc(response = df_test_final$has_exacerbation, predictor = df_test_final$pred_prob_de_novo)

# prior exac and de_novo score
model_exac_de_novo <- glm(has_exacerbation ~ de_novo_score + Exacerbation_Frequency_P2 + offset(log(total_days_since_P2/365)), data = df_test_final, family = binomial())
df_test_final$pred_prob_exac_de_novo <- predict(model_exac_de_novo, type = "response")
roc_exac_de_novo <- roc(response  = df_test_final$has_exacerbation, predictor = df_test_final$pred_prob_exac_de_novo)
```

```{r}
formula_str <- paste0("has_exacerbation ~ ", paste0(c("de_novo_score", "Exacerbation_Frequency_P2", "ATS_PackYears_P2", "SmokCigNow_P2", "race", "offset(log(total_days_since_P2/365))"), collapse = " + "))
model_de_novo_exac_py_smk_race <- glm(as.formula(formula_str), data = df_test_final, family = binomial())

beta_all <- coef(model_de_novo_exac_py_smk_race)
ci_all <- confint(model_de_novo_exac_py_smk_race, level = 0.95)
pval_all <- summary(model_de_novo_exac_py_smk_race)$coefficients[, 4]

de_novo_exac_py_smk_race_table <- data.frame(
  predictor = names(beta_all),
  beta = beta_all,
  ci_lower = ci_all[, 1],
  ci_upper = ci_all[, 2],
  pval = pval_all,
  row.names = NULL) %>%
  mutate(or = exp(beta), ci_lower = exp(ci_lower), ci_upper = exp(ci_upper), or_ci = sprintf("%.3f (%.3f, %.3f)", or, ci_lower, ci_upper)) %>%
  mutate_if(is.numeric, signif, digits=3) %>%
  filter(!predictor %in% c("(Intercept)")) %>%
  mutate(predictor = c("De novo score", "Prior exacerbation", "Pack years", "Current smoking", "Race")) %>%
  dplyr::select(predictor, or_ci, pval)

write.table(de_novo_exac_py_smk_race_table, "../data/multivariable_model_sumtab_cg_test_de_novo_adj_smk_race.txt", row.names = F, quote = F, sep = "\t")
# ChanRev26 
DT::datatable(de_novo_exac_py_smk_race_table, rownames = F)
```

```{r}
# nerd vs. de novo
roc.test(roc_nerd, roc_de_novo, method = "delong")

# nerd + exac vs. de novo + exac
roc.test(roc_nerd_exac, roc_exac_de_novo, method = "delong")
```


```{r fig.height=5, fig.width=5.5}
models <- c("De Novo", "Prior Exacerbations",  "Prior Exacerbations + De Novo")
rocs <- list(roc_de_novo, roc_exac, roc_exac_de_novo)

auc_df_de_novo <- data.frame(model = models, auc = sapply(rocs, auc))

ci_mat <- t(sapply(rocs, ci.auc))
auc_df_de_novo$ci_low  <- ci_mat[,1]
auc_df_de_novo$ci_high <- ci_mat[,3]

auc_df_de_novo$model <- factor(auc_df_de_novo$model, levels = models)


p <- ggplot(auc_df_de_novo, aes(x = model, y = auc)) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high, color = "COPDGene"), width = 0.15, linewidth = 0.8) +
  geom_segment(aes(x = as.numeric(model) - 0.075, xend = as.numeric(model) + 0.075,
                   y = auc, yend = auc, color = "COPDGene"), linewidth = 0.8) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
  scale_x_discrete(labels = c(
    "De Novo" = "De Novo Score",
    "Prior Exacerbations" = "Prior Exacerbations",
    "Prior Exacerbations + De Novo" = "Prior Exacerbations\n+ De Novo Score"
  )) +
  labs(y = "AUC (95% CI)", x = NULL, color = "Cohort") +
  scale_color_manual(values = c("COPDGene" = "#0033FF")) +
  theme_bw(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# ChanRev27
# added COPDGene legend
print(p)
ggsave("../figures_final/roc_de_novo_vs_exac_forest_vertical.png", width = 5.5, height = 5, dpi = 600)
ggsave("../figures_final/roc_de_novo_vs_exac_forest_vertical.eps", width = 5.5, height = 5, device = "eps")
```

## ntile plot
```{r warning=FALSE, message=FALSE}
df_test_final <- df_test_final %>%
  mutate(score_ntile = ntile(nerd_score, 4))

ntile_cg <- get_ntile_summary(df = df_test_final, outcome_name = "has_exacerbation", score = "nerd_score", covars = NULL, adjustment = "offset(log(total_days_since_P2/365))", model_type = "binomial", ntile = 4)
ntile_cg$score_type <- "COPDGene"

ntile_ec <- read.table("../data/quartile_results_eclipse.txt", sep = "\t", header = T)

ntile_nerd <- bind_rows(ntile_cg, ntile_ec)

ntile_nerd$x_base <- as.numeric(factor(ntile_nerd$ntile))
offset_width <- 0.1

ntile_nerd <- ntile_nerd %>%
  mutate(
    offset = case_when(
      score_type == "COPDGene" ~ -offset_width,
      score_type == "ECLIPSE" ~ offset_width
    ),
    x_pos = x_base + offset
  )
# ChanRev28
DT::datatable(ntile_nerd %>% mutate_if(is.numeric, signif, digits=3), rownames = F, options = list(scrollX=TRUE))
```


```{r fig.height=6, fig.width=4.5}
p <- ggplot(ntile_nerd, aes(x = x_pos, y = OR, color = score_type)) +

  geom_errorbar(
    aes(ymin = CI_lower, ymax = CI_upper),
    width = 0.1, linewidth = 0.8) +

  geom_segment(
    aes(x = x_pos - 0.08, xend = x_pos + 0.08, y = OR, yend = OR),
    linewidth = 1) +

  geom_hline(yintercept = 1, linetype = "dashed") +

  scale_y_log10(
    breaks = c(0.75, 1, 1.5, 2, 2.5, 3, 4, 5),
    labels = c("0.75", "1", "1.5", "2", "2.5", "3", "4", "5")
  ) +
  
  scale_x_continuous(
    breaks = unique(ntile_nerd$x_base),
    labels = sort(unique(ntile_nerd$ntile))
  ) +

  labs(
    x = "NERD Quartile",
    y = "Exacerbation Odds Ratio",
    color = "Predictor") +
  
  scale_color_manual(
    values = c(
      "COPDGene" = "#0033FF",
      "ECLIPSE" = "#00CCFF")) +

  theme_bw(base_size = 12) +
  coord_flip() +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.98, 0.02),
    legend.justification = c(1, 0),
    legend.background = element_rect(fill = "white", color = "grey70")
  )

p <- p +
  geom_text(
    aes(x = x_pos, y = OR * 1.2, label = paste0("n=", N)),
    position = position_nudge(x = 0.06), 
    size = 3.2, color = "black",
    show.legend = FALSE
  )
print(p)

ggsave("../figures_final/quartiles_nerd_cg_ec_log_rotated.png", width = 4.5, height = 6, dpi = 600)
ggsave("../figures_final/quartiles_nerd_cg_ec_log_rotated.eps", width = 4.5, height = 6, device = "eps")
```

## PPV NPV
```{r}
df_test_final$pred_pos <- df_test_final$score_ntile == 4

summary_full <- df_test_final %>%
  summarise(
    TP = sum(pred_pos & has_exacerbation == 1, na.rm = TRUE),
    FP = sum(pred_pos & has_exacerbation == 0, na.rm = TRUE),
    FN = sum(!pred_pos & has_exacerbation == 1, na.rm = TRUE),
    TN = sum(!pred_pos & has_exacerbation == 0, na.rm = TRUE),
    sens = TP / (TP + FN),
    spec = TN / (TN + FP),
    prev = mean(has_exacerbation == 1, na.rm = TRUE),
    PPV = (prev * sens) / ((prev * sens) + ((1 - prev) * (1 - spec))),
    NPV = ((1 - prev) * spec) / (((1 - prev) * spec) + (prev * (1 - sens)))
  )

summary_full$cohort <- "COPDGene"

summary_ec <- read.table("../data/ppv_npv_sumtab_ec.txt", sep = "\t", header = T)
summary_ec$cohort <- "ECLIPSE"
summary_cg_ec <- rbind(summary_full, summary_ec) %>%
  relocate(cohort, .before = TP)

write.table(summary_cg_ec, "../data/ppv_npv_sumtab_cg_ec.txt", row.names = F, quote = F, sep = "\t")
# ChanRev38
DT::datatable(summary_cg_ec, options = list(scrollX = T))
```

## violin plot
```{r fig.height=5, fig.width=6}

p <- ggplot(df_test_final, aes(x = factor(has_exacerbation), y = nerd_score)) +
  geom_violin(trim = FALSE, fill = "skyblue", alpha = 0.4) +
  geom_boxplot(width = 0.1, outlier.shape = NA, alpha = 0.7) +
  theme_minimal() +
  labs(x = "Has exacerbation", y = "NERD score") +
  stat_compare_means(method = "t.test", label.x = 1.3)
# ChanRev18
print(p)
ggsave("../figures_final/module_score_violin.png", plot = p, width = 6, height = 5, dpi = 600)
ggsave("../figures_final/module_score_violin.eps", plot = p, width = 6, height = 5, device = "eps")

```

# survival analysis 
## cox
```{r}
# 11 people have missing survival data and no excaerbation -> NA in composite event 
df_test_final %>% count(mortality_survival_vital_status, has_exacerbation)
```

```{r}
nerd_cox <- call_coxph(df_test_final, time = "composite_time", event = "composite_event", predictor = "nerd_score", adjustment_vars=NULL)
nerd_exac_cox <- call_coxph(df_test_final, time = "composite_time", event = "composite_event", predictor = "nerd_score", adjustment_vars="Exacerbation_Frequency_P2")
exac_cox <- call_coxph(df_test_final, time = "composite_time", event = "composite_event", predictor = "Exacerbation_Frequency_P2", adjustment_vars=NULL)
# score, pack years, prior exacerbation
nerd_exac_py_cox <- call_coxph(df_test_final, time = "composite_time", event = "composite_event", predictor = "nerd_score", adjustment_vars=c("Exacerbation_Frequency_P2", "ATS_PackYears_P2"))

has_exac_cox_sumtab <- rbind(nerd_cox$result, nerd_exac_cox$result, exac_cox$result, nerd_exac_py_cox$result) %>%
  dplyr::select(, predictor, n_samples, n_events, hr_ci, pval, ph_pval, c_stat, c_stat_se) %>%  
  mutate_if(is.numeric, signif, digits=3) %>%
  mutate(model = c("NERD score only", "NERD score + prior exacerbations", 
                   "Prior exacerabtions only", "NERD score + prior exacerbations + pack years"),
         predictor = c("NERD score", "NERD score", "Prior exacerabtions", "NERD score")) %>%
  relocate(model, .before = predictor)
rownames(has_exac_cox_sumtab) <- NULL

DT::datatable(has_exac_cox_sumtab, options=list(scrollX=T))
write.table(has_exac_cox_sumtab, "../data/survival_sumtab_has_exac_cox_ph.txt", row.names = F, quote = F, sep = "\t")
```

```{r}
plot(nerd_cox$ph_test, var="nerd_score", main="Schoenfeld residual")
plot(nerd_exac_cox$ph_test, var="nerd_score", main="Schoenfeld residual")
plot(exac_cox$ph_test, var="Exacerbation_Frequency_P2", main="Schoenfeld residual")
```

```{r}
# score, pack years, prior exacerbation, smk status, race
nerd_exac_py_smk_race_cox <- call_coxph(df_test_final, time = "composite_time", event = "composite_event", predictor = "nerd_score", 
                                 adjustment_vars = c("Exacerbation_Frequency_P2", "ATS_PackYears_P2", "SmokCigNow_P2", "race"))

nerd_exac_py_smk_race_cox_out <- summary(nerd_exac_py_smk_race_cox$model)$coefficients
nerd_exac_py_smk_race_cox_ci <- confint(nerd_exac_py_smk_race_cox$model)
nerd_exac_py_smk_race_cox_sumtab_cg <- data.frame(
  predictor = rownames(nerd_exac_py_smk_race_cox_out),
  hr = exp(nerd_exac_py_smk_race_cox_out[, "coef"]),
  ci_lower = exp(nerd_exac_py_smk_race_cox_ci[, 1]),
  ci_upper = exp(nerd_exac_py_smk_race_cox_ci[, 2]),
  cox_hr_ci_COPDGene = sprintf("%.2f (%.2f, %.2f)", exp(nerd_exac_py_smk_race_cox_out[, "coef"]), exp(nerd_exac_py_smk_race_cox_ci[, 1]), exp(nerd_exac_py_smk_race_cox_ci[, 2])),
  cox_pval_COPDGene = nerd_exac_py_smk_race_cox_out[, "Pr(>|z|)"],
  row.names = NULL
) %>%
  mutate(cox_pval_COPDGene = ifelse(cox_pval_COPDGene < 0.001, "<.001",
                ifelse(cox_pval_COPDGene < 0.01, sub("^0", "", sprintf("%.3f", cox_pval_COPDGene)),
                                    sub("^0", "", sprintf("%.2f", cox_pval_COPDGene))))) %>%  
  mutate(predictor = c("NERD score", "Prior exacerbation", "Pack years", "Current smoking", "Race")) %>%
  dplyr::select(predictor, cox_hr_ci_COPDGene, cox_pval_COPDGene)
DT::datatable(nerd_exac_py_smk_race_cox_sumtab_cg)
```

```{r}
nerd_exac_py_smk_race_cox$ph_test

plot(nerd_exac_py_smk_race_cox$ph_test, var="nerd_score", main="Schoenfeld residual")
plot(nerd_exac_py_smk_race_cox$ph_test, var="Exacerbation_Frequency_P2", main="Schoenfeld residual")
```

## competing risk
```{r}
df_test_final <- df_test_final %>%
  mutate(
    event_type = case_when(
      is.na(composite_event) ~ NA_integer_,
      died & has_exac & total_days_since_P2 <= time_to_first_exac ~ 2L,
      died & has_exac ~ 1L,
      died %in% TRUE ~ 2L,      # use in to avoid na 
      has_exac %in% TRUE ~ 1L,
      TRUE ~ 0L
    )
  )
# sanity check
table(df_test_final$event_type, useNA = "ifany")
table(df_test_final$composite_event, useNA = "ifany")
```

```{r}
library(cmprsk)
cov_matrix <- model.matrix(
  ~ nerd_score,
  data = df_test_final
)[, -1, drop = FALSE]

crr_fit_exac <- crr(
  ftime = df_test_final$composite_time,
  fstatus = df_test_final$event_type,
  cov1 = cov_matrix,
  failcode = 1,   # exacerbation
  cencode = 0
)

summary(crr_fit_exac)
```

## km
```{r fig.height=6, fig.width=4.5}
km_plot_df <- df_test_final %>%
  dplyr::select(composite_time, composite_event, nerd_score, total_days_since_P2) %>%
  mutate(
    nerd_quartile = cut(nerd_score, 
                        breaks = quantile(nerd_score, probs = seq(0, 1, 0.25), na.rm = TRUE),
                        include.lowest = TRUE, labels = c("Q1", "Q2", "Q3", "Q4")))

km_plot_df_Q1_Q4 <- km_plot_df %>% 
  filter(nerd_quartile %in% c("Q1", "Q4")) %>%
  mutate(nerd_quartile = droplevels(nerd_quartile))

# coin for Fleming-Harrington test
# pvalue for Q1 vs Q4
fh_test <- logrank_test(Surv(composite_time, composite_event) ~ nerd_quartile, data = km_plot_df_Q1_Q4, type = "Fleming-Harrington", rho = 0, gamma = 1)

km_fit_all <- survfit(
  Surv(composite_time, composite_event) ~ nerd_quartile,
  data = km_plot_df
)

time_by_q <- km_plot_df %>%
  group_by(nerd_quartile) %>%
  summarise(
    time_to_event_iqr = sprintf("%.0f [%.0f, %.0f]", median(composite_time, na.rm = TRUE), 
                        quantile(composite_time, 0.25, na.rm = TRUE), 
                        quantile(composite_time, 0.75, na.rm = TRUE)),
    time_followed_iqr = sprintf("%.0f [%.0f, %.0f]", median(total_days_since_P2, na.rm = TRUE), 
                        quantile(total_days_since_P2, 0.25, na.rm = TRUE), 
                        quantile(total_days_since_P2, 0.75, na.rm = TRUE))) %>%
  mutate(cohort = "COPDGene") %>%
  relocate(cohort)

p <- ggsurvplot(
  km_fit_all,
  data = km_plot_df,
  censor = TRUE,
  xlab = "Days",
  ylab = "Exacerbation-Free Survival",
  title = "COPDGene",
  legend.title = "NERD Quartile",
  legend.labs = c("Q1", "Q2", "Q3", "Q4"),
  risk.table = TRUE,
  risk.table.title = "Number Event-Free",
  tables.theme = theme_survminer() + 
    theme(plot.title = element_text(size = 12)),
  palette = c("#FFAA00", "#FF6600", "#CC0000", "#660000")
)

p$plot <- p$plot +
  annotate("text", x = 100, y = 0.1, hjust = 0, label = sprintf("p(Q1 vs. Q4) = %.3f", pvalue(fh_test)), size = 5)

print(p)
png("../figures_final/kaplan_meier_all_q.png", width = 4.5, height = 6, units = "in", res = 600)
print(p)
dev.off()
cairo_ps("../figures_final/kaplan_meier_all_q.eps", width = 4.5, height = 6)
print(p)
dev.off()
```

# stratified on prior exacerbation
## AUC compare
```{r}
df_test_zero_exac <- df_test_final %>% filter(Exacerbation_Frequency_P2 == 0)
df_test_prior_exac <- df_test_final %>% filter(Exacerbation_Frequency_P2 > 0)

# nerd only
formula_str <- paste0("has_exacerbation ~ ", "nerd_score", "+ offset(log(total_days_since_P2/365))")
model_nerd_exac1 <- glm(as.formula(formula_str), data = df_test_prior_exac, family = binomial())
model_nerd_exac0 <- glm(as.formula(formula_str), data = df_test_zero_exac, family = binomial())
roc_nerd_exac1 <- roc(df_test_prior_exac$has_exacerbation, predict(model_nerd_exac1, type = "response"))
roc_nerd_exac0 <- roc(df_test_zero_exac$has_exacerbation, predict(model_nerd_exac0, type = "response"))
```


```{r fig.height=5, fig.width=6}
models <- c("NERD score no prior exacerbation", "NERD score with prior exacerbation")
rocs <- list(roc_nerd_exac0, roc_nerd_exac1)

auc_df <- data.frame(model = models, auc = sapply(rocs, auc))

ci_mat <- t(sapply(rocs, ci.auc))
auc_df$ci_low  <- ci_mat[,1]
auc_df$ci_high <- ci_mat[,3]

auc_df_ec <- read.table("../data/auc_df_nerd_vs_smk_prior_exac_strat_eclipse.txt", header = T, sep = "\t")
auc_df_merged <- merge(auc_df, auc_df_ec, by = "model", suffixes = c("_COPDGene", "_ECLIPSE"))
auc_df_merged$model <- factor(auc_df_merged$model, levels = models)

auc_long <- auc_df_merged %>%
  dplyr::select(model, auc_COPDGene, ci_low_COPDGene, ci_high_COPDGene,
         auc_ECLIPSE,  ci_low_ECLIPSE,  ci_high_ECLIPSE) %>%
  pivot_longer(cols = -model, names_to = c(".value", "cohort"), names_pattern = "(.+)_(.+)$") %>% 
  mutate(model_num = as.numeric(model), offset = ifelse(cohort == "COPDGene", -0.2, 0.2), x_pos = model_num + offset)

p <- ggplot(auc_long, aes(y = auc, color = cohort)) +

  geom_errorbar(aes(x = x_pos, ymin = ci_low, ymax = ci_high), width = 0.1, linewidth = 0.8) +
  geom_segment(aes(x = x_pos - 0.05, xend = x_pos + 0.05, y = auc, yend = auc), linewidth = 0.8) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
  
  scale_x_continuous(breaks = unique(auc_long$model_num), expand = expansion(add = 0.4),
                     labels = c("No Prior Exacerbations", "With Prior Exacerbations")) +
  labs(y = "AUC NERD Score (95% CI)", x = NULL, color = "Cohort") +

  theme_bw(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  scale_color_manual(values = c("COPDGene" = "#0033FF",
                                "ECLIPSE" = "#00CCFF"))

print(p)
ggsave("../figures_final/roc_nerd_exac_forest_vertical_exac0_vs_exac1.png", width = 6, height = 5, dpi = 600)
ggsave("../figures_final/roc_nerd_exac_forest_vertical_exac0_vs_exac1.eps", width = 6, height = 5, device = "eps")

```

```{r fig.height=5, fig.width=7}

p <- ggplot(df_test_prior_exac, aes(x = factor(has_exacerbation), y = nerd_score)) +
  geom_violin(trim = FALSE, fill = "skyblue", alpha = 0.4) +
  geom_boxplot(width = 0.1, outlier.shape = NA, alpha = 0.7) +
  theme_minimal() +
  labs(x = "Has exacerbation", y = "NERD score", title = "Stratified cohort with 1 or more prior exacerbations") +
  stat_compare_means(method = "t.test", label.x = 1.3)

print(p)

p <- ggplot(df_test_zero_exac, aes(x = factor(has_exacerbation), y = nerd_score)) +
  geom_violin(trim = FALSE, fill = "skyblue", alpha = 0.4) +
  geom_boxplot(width = 0.1, outlier.shape = NA, alpha = 0.7) +
  theme_minimal() +
  labs(x = "Has exacerbation", y = "NERD score", title = "Stratified cohort with 0 prior exacerbations") +
  stat_compare_means(method = "t.test", label.x = 1.3)

print(p)
```

## no prior
```{r}
ntile_cg_exac0 <- get_ntile_summary(df = df_test_zero_exac, outcome_name = "has_exacerbation", score = "nerd_score", covars = NULL, adjustment = "offset(log(total_days_since_P2/365))", model_type = "binomial", ntile = 4)
ntile_cg_exac0$score_type <- "COPDGene"

ntile_ec_exac0 <- read.table("../data/quartile_results_eclipse_0_prior_exac.txt", sep = "\t", header = T)

ntile_nerd_exac0 <- bind_rows(ntile_cg_exac0, ntile_ec_exac0)

ntile_nerd_exac0$x_base <- as.numeric(factor(ntile_nerd_exac0$ntile))
offset_width <- 0.1

ntile_nerd_exac0 <- ntile_nerd_exac0 %>%
  mutate(
    offset = case_when(
      score_type == "COPDGene" ~ -offset_width,
      score_type == "ECLIPSE" ~ offset_width
    ),
    x_pos = x_base + offset
  )

# ChanRev33
DT::datatable(ntile_nerd_exac0 %>% mutate_if(is.numeric, signif, digits=3), rownames = F, options = list(scrollX=TRUE))
```



```{r fig.height=6, fig.width=4.5}
p <- ggplot(ntile_nerd_exac0, aes(x = x_pos, y = OR, color = score_type)) +

  geom_errorbar(
    aes(ymin = CI_lower, ymax = CI_upper),
    width = 0.1, linewidth = 0.8) +

  geom_segment(
    aes(x = x_pos - 0.08, xend = x_pos + 0.08, y = OR, yend = OR),
    linewidth = 1) +

  geom_hline(yintercept = 1, linetype = "dashed") +

  scale_y_log10(
    breaks = c(0.75, 1, 1.5, 2, 3, 4, 6),
    labels = c("0.75", "1", "1.5", "2", "3", "4", "6")
  ) +
  scale_x_continuous(
    breaks = unique(ntile_nerd_exac0$x_base),
    labels = sort(unique(ntile_nerd_exac0$ntile))
  ) +

  labs(
    x = "NERD Quartile",
    y = "Exacerbation Odds Ratio",
    # title = "Stratified cohort with 0 prior exacerbation", 
    color = "Predictor") +
  
  scale_color_manual(
    values = c(
      "COPDGene" = "#0033FF",
      "ECLIPSE" = "#00CCFF"),
    labels = c(
      "COPDGene" = "COPDGene\nNo Prior Exacerbations",
      "ECLIPSE"  = "ECLIPSE\nNo Prior Exacerbations")) +

  theme_bw(base_size = 12) +
  theme(legend.spacing.y = unit(0.5, "cm"), legend.key.height = unit(1, "cm")) + 
  coord_flip() + 
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.98, 0.02),
    legend.justification = c(1, 0),
    legend.background = element_rect(fill = "white", color = "grey70")
  )
  
p <- p +
  geom_text(
    aes(x = x_pos, y = OR * 1.2, label = paste0("n=", N)),
    position = position_nudge(x = 0.06), 
    size = 3.2, color = "black",
    show.legend = FALSE
  )
print(p)

ggsave("../figures_final/quartiles_nerd_cg_ec_no_prior_exacerbation_log_rotated.png", width = 4.5, height = 6, dpi = 600)
ggsave("../figures_final/quartiles_nerd_cg_ec_no_prior_exacerbation_log_rotated.eps", width = 4.5, height = 6, device = "eps")

```

```{r}
# no longer adjusting for prior exac
nerd_cox_exac0 <- call_coxph(df_test_zero_exac, time = "composite_time", event = "composite_event", predictor = "nerd_score", adjustment_vars=NULL)

has_exac_cox_sumtab_exac0 <- rbind(nerd_cox_exac0$result) %>%
  mutate(outcome = "has_exacerbation")%>% 
  dplyr::select(outcome, predictor, n_samples, n_events, hr_ci, pval, ph_pval, c_stat, c_stat_se) %>%  
  mutate_if(is.numeric, signif, digits=3) %>%
  mutate(model = c("NERD score only"),
         predictor = c("NERD score")) %>%
  relocate(model, .before = predictor) %>%
  dplyr::select(-outcome)
rownames(has_exac_cox_sumtab_exac0) <- NULL

DT::datatable(has_exac_cox_sumtab_exac0, options=list(scrollX=T))
write.table(has_exac_cox_sumtab_exac0, "../data/survival_sumtab_has_exac_no_exac_cox_ph.txt", row.names = F, quote = F, sep = "\t")
```

```{r}
plot(nerd_cox_exac0$ph_test, var="nerd_score", main="Schoenfeld residual")
```


```{r fig.height=6, fig.width=4.5}
km_plot_df_exac0 <- df_test_zero_exac %>%
  dplyr::select(composite_time, composite_event, nerd_score, total_days_since_P2) %>%
  mutate(
    nerd_quartile = cut(nerd_score, 
                        breaks = quantile(nerd_score, probs = seq(0, 1, 0.25), na.rm = TRUE),
                        include.lowest = TRUE, labels = c("Q1", "Q2", "Q3", "Q4")))

km_plot_df_Q1_Q4_exac0 <- km_plot_df_exac0 %>% 
  filter(nerd_quartile %in% c("Q1", "Q4")) %>%
  mutate(nerd_quartile = droplevels(nerd_quartile))

fh_test_exac0 <- logrank_test(Surv(composite_time, composite_event) ~ nerd_quartile, data = km_plot_df_Q1_Q4_exac0, type = "Fleming-Harrington", rho = 0, gamma = 1)

km_fit_exac0_all <- survfit(
  Surv(composite_time, composite_event) ~ nerd_quartile,
  data = km_plot_df_exac0
)

time_by_q_exac0 <- km_plot_df_exac0 %>%
  group_by(nerd_quartile) %>%
  summarise(
    time_to_event_iqr = sprintf("%.0f [%.0f, %.0f]", median(composite_time, na.rm = TRUE), 
                        quantile(composite_time, 0.25, na.rm = TRUE), 
                        quantile(composite_time, 0.75, na.rm = TRUE)),
    time_followed_iqr = sprintf("%.0f [%.0f, %.0f]", median(total_days_since_P2, na.rm = TRUE), 
                        quantile(total_days_since_P2, 0.25, na.rm = TRUE), 
                        quantile(total_days_since_P2, 0.75, na.rm = TRUE))) %>%
  mutate(cohort = "COPDGene no prior exacerbations") %>%
  relocate(cohort)

p <- ggsurvplot(
  km_fit_exac0_all,
  data = km_plot_df_exac0,
  censor = TRUE,
  xlab = "Days",
  ylab = "Exacerbation-Free Survival",
  title = "COPDGene No Prior Exacerbations",
  legend.title = "NERD Quartile",
  legend.labs = c("Q1", "Q2", "Q3", "Q4"),
  risk.table = TRUE,
  risk.table.title = "Number Event-Free",
  tables.theme = theme_survminer() + 
    theme(plot.title = element_text(size = 12)),
  palette = c("#FFAA00", "#FF6600", "#CC0000", "#660000")
)
p$plot <- p$plot +
  annotate("text", x = 100, y = 0.1, hjust = 0, label = sprintf("p(Q1 vs. Q4) = %.3f", pvalue(fh_test_exac0)), size = 5)

print(p)
png("../figures_final/kaplan_meier_all_q_zero_prior_exac.png", width = 4.5, height = 6, units = "in", res = 600)
print(p)
dev.off()
cairo_ps("../figures_final/kaplan_meier_all_q_zero_prior_exac.eps", width = 4.5, height = 6)
print(p)
dev.off()
```

```{r}

df_test_zero_exac$pred_pos <- df_test_zero_exac$score_ntile == 4

summary_full <- df_test_zero_exac %>%
  summarise(
    TP = sum(pred_pos & has_exacerbation == 1, na.rm = TRUE),
    FP = sum(pred_pos & has_exacerbation == 0, na.rm = TRUE),
    FN = sum(!pred_pos & has_exacerbation == 1, na.rm = TRUE),
    TN = sum(!pred_pos & has_exacerbation == 0, na.rm = TRUE),
    sens = TP / (TP + FN),
    spec = TN / (TN + FP),
    prev = mean(has_exacerbation == 1, na.rm = TRUE),
    PPV = (prev * sens) / ((prev * sens) + ((1 - prev) * (1 - spec))),
    NPV = ((1 - prev) * spec) / (((1 - prev) * spec) + (prev * (1 - sens)))
  )

summary_full
```


## with prior
```{r}
ntile_cg_exac1 <- get_ntile_summary(df = df_test_prior_exac, outcome_name = "has_exacerbation", score = "nerd_score", covars = NULL, adjustment = "offset(log(total_days_since_P2/365))", model_type = "binomial", ntile = 4)
ntile_cg_exac1$score_type <- "COPDGene"

ntile_ec_exac1 <- read.table("../data/quartile_results_eclipse_with_prior_exac.txt", sep = "\t", header = T)

ntile_nerd_exac1 <- bind_rows(ntile_cg_exac1, ntile_ec_exac1)

ntile_nerd_exac1$x_base <- as.numeric(factor(ntile_nerd_exac1$ntile))
offset_width <- 0.1

ntile_nerd_exac1 <- ntile_nerd_exac1 %>%
  mutate(
    offset = case_when(
      score_type == "COPDGene" ~ -offset_width,
      score_type == "ECLIPSE" ~ offset_width
    ),
    x_pos = x_base + offset
  )

DT::datatable(ntile_nerd_exac1 %>% mutate_if(is.numeric, signif, digits=3), rownames = F, options = list(scrollX=TRUE))
```


```{r fig.height=6, fig.width=4.5}
p <- ggplot(ntile_nerd_exac1, aes(x = x_pos, y = OR, color = score_type)) +

  geom_errorbar(
    aes(ymin = CI_lower, ymax = CI_upper),
    width = 0.1, linewidth = 0.8) +

  geom_segment(
    aes(x = x_pos - 0.08, xend = x_pos + 0.08, y = OR, yend = OR),
    linewidth = 1) +

  geom_hline(yintercept = 1, linetype = "dashed") +

  scale_y_log10(
    breaks = c(0.5, 1, 2, 3, 5, 8),
    labels = c("0.5", "1", "2", "3", "5", "8")
  ) +
  
  scale_x_continuous(
    breaks = unique(ntile_nerd_exac1$x_base),
    labels = sort(unique(ntile_nerd_exac1$ntile))
  ) +

  labs(
    x = "NERD Quartile",
    y = "Exacerbation Odds Ratio",
    # title = "Stratified cohort with prior exacerbation", 
    color = "Predictor") +
  
  scale_color_manual(
    values = c(
      "COPDGene" = "#0033FF",
      "ECLIPSE" = "#00CCFF"),
    labels = c(
      "COPDGene" = "COPDGene\nWith Prior Exacerbations",
      "ECLIPSE"  = "ECLIPSE\nWith Prior Exacerbations")) +

  theme_bw(base_size = 12) +
  theme(legend.spacing.y = unit(0.5, "cm"), legend.key.height = unit(1, "cm")) +
  coord_flip() + 
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.98, 0.02),
    legend.justification = c(1, 0),
    legend.background = element_rect(fill = "white", color = "grey70")
  )
  
p <- p +
  geom_text(
    aes(x = x_pos, y = OR * 1.35, label = paste0("n=", N)),
    position = position_nudge(x = 0.06), 
    size = 3.2, color = "black",
    show.legend = FALSE
  )
print(p)

ggsave("../figures_final/quartiles_nerd_cg_ec_with_prior_exacerbation_log_rotated.png", width = 4.5, height = 6, dpi = 600)
ggsave("../figures_final/quartiles_nerd_cg_ec_with_prior_exacerbation_log_rotated.eps", width = 4.5, height = 6, device = "eps")

```

```{r}
# no longer adjusting for prior exac
nerd_cox_exac1 <- call_coxph(df_test_prior_exac, time = "composite_time", event = "composite_event", predictor = "nerd_score", adjustment_vars=NULL)

has_exac_cox_sumtab_exac1 <- rbind(nerd_cox_exac1$result) %>%
  mutate(outcome = "has_exacerbation")%>% 
  dplyr::select(outcome, predictor, n_samples, n_events, hr_ci, pval, ph_pval, c_stat, c_stat_se) %>%  
  mutate_if(is.numeric, signif, digits=3) %>%
  mutate(model = c("NERD score only"),
         predictor = c("NERD score")) %>%
  relocate(model, .before = predictor) %>%
  dplyr::select(-outcome)
rownames(has_exac_cox_sumtab_exac1) <- NULL

DT::datatable(has_exac_cox_sumtab_exac1, options=list(scrollX=T))
write.table(has_exac_cox_sumtab_exac1, "../data/survival_sumtab_has_exac_with_exac_cox_ph.txt", row.names = F, quote = F, sep = "\t")
```

```{r}
plot(nerd_cox_exac1$ph_test, var="nerd_score", main="Schoenfeld residual")
```


```{r fig.height=6, fig.width=4.5}
km_plot_df_exac1 <- df_test_prior_exac %>%
  dplyr::select(composite_time, composite_event, nerd_score, total_days_since_P2) %>%
  mutate(
    nerd_quartile = cut(nerd_score, 
                        breaks = quantile(nerd_score, probs = seq(0, 1, 0.25), na.rm = TRUE),
                        include.lowest = TRUE, labels = c("Q1", "Q2", "Q3", "Q4")))

km_plot_df_Q1_Q4_exac1 <- km_plot_df_exac1 %>% 
  filter(nerd_quartile %in% c("Q1", "Q4")) %>%
  mutate(nerd_quartile = droplevels(nerd_quartile))

fh_test_exac1 <- logrank_test(Surv(composite_time, composite_event) ~ nerd_quartile, data = km_plot_df_Q1_Q4_exac1, type = "Fleming-Harrington", rho = 0, gamma = 1)

km_fit_exac1_all <- survfit(
  Surv(composite_time, composite_event) ~ nerd_quartile,
  data = km_plot_df_exac1
)

time_by_q_exac1 <- km_plot_df_exac1 %>%
  group_by(nerd_quartile) %>%
  summarise(
    time_to_event_iqr = sprintf("%.0f [%.0f, %.0f]", median(composite_time, na.rm = TRUE), 
                        quantile(composite_time, 0.25, na.rm = TRUE), 
                        quantile(composite_time, 0.75, na.rm = TRUE)),
    time_followed_iqr = sprintf("%.0f [%.0f, %.0f]", median(total_days_since_P2, na.rm = TRUE), 
                        quantile(total_days_since_P2, 0.25, na.rm = TRUE), 
                        quantile(total_days_since_P2, 0.75, na.rm = TRUE))) %>%
  mutate(cohort = "COPDGene with prior exacerbations") %>%
  relocate(cohort)

p <- ggsurvplot(
  km_fit_exac1_all,
  data = km_plot_df_exac1,
  censor = TRUE,
  xlab = "Days",
  ylab = "Exacerbation-Free Survival",
  title = "COPDGene With Prior Exacerbations",
  legend.title = "NERD Quartile",
  legend.labs = c("Q1", "Q2", "Q3", "Q4"),
  risk.table = TRUE,
  risk.table.title = "Number Event-Free",
  tables.theme = theme_survminer() + 
    theme(plot.title = element_text(size = 12)),
  palette = c("#FFAA00", "#FF6600", "#CC0000", "#660000")
)
p$plot <- p$plot +
  annotate("text", x = 100, y = 0.1, hjust = 0, label = sprintf("p(Q1 vs. Q4) = %.3f", pvalue(fh_test_exac1)), size = 5)
p$plot <- p$plot + theme(plot.title = element_text(hjust = 0.7))

print(p)
png("../figures_final/kaplan_meier_all_q_with_prior_exac.png", width = 4.5, height = 6, units = "in", res = 600)
print(p)
dev.off()
cairo_ps("../figures_final/kaplan_meier_all_q_with_prior_exac.eps", width = 4.5, height = 6)
print(p)
dev.off()
```

# sumtabs
## n subjects and events
```{r}
n_exac_sumtab_cg <- tibble(
  cohort = c("COPDGene", "COPDGene no prior exacerbations", "COPDGene with prior exacerbations"),
  n = c(nrow(df_test_final), nrow(df_test_prior_exac), nrow(df_test_zero_exac)),
  n_has_exacerbation = c(sum(as.integer(as.character(df_test_final$has_exacerbation)), na.rm = TRUE),
                          sum(as.integer(as.character(df_test_prior_exac$has_exacerbation)), na.rm = TRUE),
                          sum(as.integer(as.character(df_test_zero_exac$has_exacerbation)), na.rm = TRUE))
)
n_exac_sumtab_ec <- read.table("../data/n_subj_n_event_EC.txt", header = T, sep = "\t")
n_exac_sumtab <- rbind(n_exac_sumtab_cg, n_exac_sumtab_ec)
DT::datatable(n_exac_sumtab)
write.table(n_exac_sumtab, "../data/n_subj_n_event.txt", row.names = F, quote = F, sep = "\t")
```

## table one additional
```{r}
cg_flat <- fread("path/to/cg_flat.txt", data.table = F)

df_test_final2 <- df_test_final %>% mutate(
    nerd_quartile = cut(nerd_score, 
                        breaks = quantile(nerd_score, probs = seq(0, 1, 0.25), na.rm = TRUE),
                        include.lowest = TRUE, labels = c("Q1", "Q2", "Q3", "Q4")))


more_vars <- c("BMI_P2", "fvc_pp_gli_global_P2", "finalGold_P2", "MMRCDyspneaScor_P2", "SGRQ_scoreTotal_P2", "BODE_P2", "Resting_SaO2_P2")
more_vars_df <- cg_flat %>%
  dplyr::select(sid, all_of(more_vars))

df_test_final2 <- df_test_final2 %>% merge(more_vars_df, all.x = T, by = "sid")

df_test_final2 <- df_test_final2 %>%
  mutate(gold_stage = case_when(
    FEV1_FVC_post_P2 < 0.7 & fev1_pp_gli_global_P2 >= 50 & fev1_pp_gli_global_P2 < 80 ~ "GOLD 2",
    FEV1_FVC_post_P2 < 0.7 & fev1_pp_gli_global_P2 >= 30 & fev1_pp_gli_global_P2 < 50 ~ "GOLD 3",
    FEV1_FVC_post_P2 < 0.7 & fev1_pp_gli_global_P2 < 30 ~ "GOLD 4",
    TRUE ~ NA_character_
  ),
  gold_stage = factor(gold_stage, levels = c("GOLD 2", "GOLD 3", "GOLD 4")))


vars_to_fix <- c("BMI_P2", "total_days_since_P2", "fev1_pp_gli_global_P2", "fvc_pp_gli_global_P2", "gold_stage", "MMRCDyspneaScor_P2", "SGRQ_scoreTotal_P2", "BODE_P2", "Resting_SaO2_P2", "Exacerbation_Frequency_P2")
table1_fix <- CreateTableOne(vars = vars_to_fix, strata = "nerd_quartile", data = df_test_final2, factorVars = "gold_stage", test = TRUE, addOverall = T)
table1_fix_df <- print(table1_fix, nonnormal = c("total_days_since_P2", "fev1_pp_gli_global_P2"), 
                       showAllLevels = TRUE, quote = FALSE, noSpaces = TRUE,
                       contDigits = 1, catDigits = 1, pDigits = 2)
write.table(table1_fix_df, "../data/table1_cg_by_q_more_vars.txt", sep = "\t", row.names = TRUE, quote = FALSE)
```

## time by IQR
```{r}
time_by_q_cg <- rbind(time_by_q, time_by_q_exac0, time_by_q_exac1)
time_by_q_ec <- read.table("../data/time_to_event_by_score_quartile_EC.txt", header = T, sep = "\t")
time_by_q <- rbind(time_by_q_cg, time_by_q_ec)
DT::datatable(time_by_q)
write.table(time_by_q, "../data/time_to_event_by_score_quartile.txt", row.names = F, quote = F, sep = "\t")
```

## cox
```{r}
has_exac_cox_sumtab_ec <- read.table("../data/survival_sumtab_has_exac_cox_ph_EC.txt", header = T, sep = "\t")
has_exac_cox_sumtab_exac0_ec <- read.table("../data/survival_sumtab_has_exac_no_prior_exac_cox_ph_EC.txt", header = T, sep = "\t")
has_exac_cox_sumtab_exac1_ec <- read.table("../data/survival_sumtab_has_exac_with_prior_exac_cox_ph_EC.txt", header = T, sep = "\t")
combined_cox_sumtab <- rbind(
  has_exac_cox_sumtab %>% filter(model == "NERD score only") %>% mutate(cohort = "COPDGene"),
  has_exac_cox_sumtab_exac0 %>% mutate(cohort = "COPDGene no prior exacerbation"),
  has_exac_cox_sumtab_exac1 %>% mutate(cohort = "COPDGene with prior exacerbation"),
  has_exac_cox_sumtab_ec %>% filter(model == "NERD score only") %>% mutate(cohort = "ECLIPSE"),
  has_exac_cox_sumtab_exac0_ec %>% mutate(cohort = "ECLIPSE no prior exacerbation"),
  has_exac_cox_sumtab_exac1_ec %>% mutate(cohort = "ECLIPSE with prior exacerbation")
) %>%
  dplyr::select(-model, -predictor) %>%
  relocate("cohort", .before = n_samples)
DT::datatable(combined_cox_sumtab)

write.table(combined_cox_sumtab, "../data/survival_sumtab_has_exac_all_strata_cox_ph_cg_ec.txt", row.names = F, quote = F, sep = "\t")
```

## cox and multivar
```{r}
nerd_exac_py_smk_race_cox_sumtab_ec <- read.table("../data/survival_sumtab_has_exac_all_vars_cox_ph_EC.txt", header = T, sep = "\t")
nerd_exac_py_smk_race_cox_sumtab <- merge(nerd_exac_py_smk_race_cox_sumtab_cg, nerd_exac_py_smk_race_cox_sumtab_ec, by = "predictor", all.x=T)

colnames(nerd_exac_py_smk_race_merged) <- c("predictor", "binomial_or_ci_COPDGene", "binomial_pval_COPDGene", "binomial_or_ci_ECLIPSE", "binomial_pval_ECLIPSE")

cox_multivar <- merge(nerd_exac_py_smk_race_merged, nerd_exac_py_smk_race_cox_sumtab, by="predictor", all.x=T)

write.table(cox_multivar, "../data/multivar_model_cox_sumtab_cg_ec.txt", row.names = F, quote = F, sep = "\t")
DT::datatable(cox_multivar, rownames = F, options = list(scrollX=TRUE))
```

# session info
```{r}
sessionInfo()
```

